---
title: "Asset Returns"
subtitle: "Transforming Prices to Returns in Financial Econometricss"
author: "Barry Quinn <br> and <br> <span class='glowinline'>Digital Dave</span>"
css: mycssblend.css
logo: img/qbslogo.png
footer: "Advanced Financial Data Analytics"
format: 
  revealjs:
    slide-number: c/t
---

```{r}
load("fin7028_dat.RData")
```

## Introduction
::: {.discussion}
In financial econometrics, analyzing returns rather than prices is driven by both theoretical and practical considerations. This approach aligns with key econometric models and economic theories, focusing on the stationary nature of returns, the mitigation of spurious regression problems, and the alignment with concepts of risk and reward.
:::

## Theoretical Reasons

1. **Stationarity**: Returns are more likely to be stationary compared to prices, making them suitable for statistical and econometric models.
2. **Non-Stationary Data Issues**: Returns reduce the risk of spurious regression problems by exhibiting weaker forms of non-stationarity.
3. **Economic Theory Alignment**: Analyzing returns aligns with financial economics theories focusing on risk and reward.
4. **Volatility Modeling**: Returns are essential for modeling financial market volatility, facilitating the use of models like GARCH.

## Transaction Costs

1. **Round Trip Transaction Costs**: These include all costs incurred in an investment cycle, affecting the net return on investments.
2. **Impact on Returns Analysis**: Analyzing returns rather than prices accounts for the effect of transaction costs on investment returns.
3. **Modeling and Risk Assessment**: Transaction costs are crucial for realistic econometric modeling of risk and return.
4. **Behavioral Implications**: Transaction costs influence trading frequency and market liquidity and volatility.

## Practical Reasons

1. **Comparability**: Returns allow for meaningful performance comparisons across different assets.
2. **Simplicity in Modeling**: Focusing on returns simplifies the complexity involved in modeling non-stationary price series.
3. **Risk Management**: Analyzing returns directly supports financial risk management strategies.
4. **Efficient Market Considerations**: Returns reflect new information in efficient markets, making them a focal point over price levels.

## One-period simple returns 

- Let _P~t~_ be the price of an asset at time index _t_ and assume the asset pays no dividend.
  - **One-Period Simple Return**:
  - Holding the asset for one period from date t-1 to date t would result in a _simple gross return_

$$ 
1+R_t= \frac{P_t}{P_{t-1}} \;or\; P_t=P_{t-1}(1+R_t)
$$

* The corresponding _simple net return_ or _simple return_ is:
    
$$
R_t= \frac{P_t}{P_{t-1}}-1= \;or\; \frac{P_t-P_{t-1}}{P_{t-1}}
$$


## Multiperiod simple returns 

- Holding the asset for k period between dates t-k and t gives a k period simple gross return:

$$1+R_t[k] = \frac{P_t}{P_{t-k}}=\frac{P_t}{P_{t-1}}\times\frac{P_{t-1}}{P_{t-2}} \times \dots \times \frac{P_{t-k+1}}{P_{t-k}}$$
$$=(1+R_t)(1+R_{t-1})\dots(1+R_{t-k+1})$$
$$=\prod^{k-1}_{j=0}(1+R_{t-j})$$

- The _k_ period simple gross return is simply the product of the _k_ one-period simple gross returns; or the _compound return_.
- The k-period simple net return is $R_t[k]= \frac{(P_t-P_{t-k})}{P_{t-k}}$


## Time interval
* In practice, the actual time interval is important in discussing and comparing returns (e.g., monthly returns or annual returns).
* If the time interval is not given, it is implicitly assumed to be one year.
* If the asset is held for _k_ years, then the annualised (average) return is defined as:

$$\text{Annualised}\{R_t[k]\} = \left[\prod^{k-1}_{j=0}(1+R_{t-j})\right]^{1/k}-1$$

## Time interval

* As one-period returns tend to be small, a first-order Taylor expansion to approximate the annualised returns is:

$$\text{Annualised}\{R_t[k]\} \approx \frac{1}{k}\sum^{k-1}_{j=0}R_{t-j}$$


* Accuracy of this approximation may not be sufficient in some applications.

## Continuously compounded returns

* The natural logarithm of the simple gross return of an asset is called the _continuously compounded return_ or _log return_

$$
r_t=ln(1+R_t)=ln \left(\frac{P_t}{P_{t-1}} \right)=p_t-p_{t-1} \
\text{where }p_t=ln(P_t)
$$

### Advantages
* Firstly, continuously compounded multiperiod return is simply the sum of the continuously compounded one-period returns.

$$r_t[k]=ln(1+R_t[k])=ln[(1+R_t)(1+R_{t-1})\dots(1+R_{t-k+1})]$$

$$=ln(1+R_t)+ln(1+R_{t-1})+\dots +ln(1+R_{t-k+1})$$

$$=r_t+r_{t-1}+\dots+r_{t-k+1}$$

* Secondly, statistical properties of log returns are **tractable**, easy to calculate with.


## Portfolio returns

* The simple net return of a portfolio consisting of _N_ assets is a weighted average of the simple net returns of the assets involved, where the weight on each asset is the percentage of the portfolio's value invested in that asset.

* Let _p_ be a portfolio that places weights _w~i~_ on asset _i_.  

* Then the simple return of _p_ at time _t_ is:

$$R_{p,t}=\sum_{i=1}^{N}w_iR_{it}\text{ where}\sum w_i=1$$


## Portfolio returns

* log returns of a portfolio do not have this additive property, but if the simple returns _R~it~_ are all small in magnitude then we have:

$$r_{p,t} \approx \sum_{i=1}^{N}w_ir_{it}$$

* $r_{p,t}$ is the continuously compounded return of the portfolio.
* This approximation is usually used to study portfolio returns.

## Dividend payment

* Asset returns must be modified if there is a periodic dividend
* Let _D~t~_ be the dividend of an asset between dates _t-1_ and _t_, the simple net return and continuously compounded return at time t become:

$$R_t= \frac{P_t+D_t}{P_{t-1}}-1 \; , \;r_t=ln(P_t+D_t)-ln(P_{t-1})$$

* This adjustment is commonly referred to as a *total return*.

## Excess return

::: columns
:::{.column}

* Excess returns of an assets at time _t_ is the difference between the asset's return and the return on some reference asset.
* The reference asset is often taken to be riskless such as a short-term UK government bond returns.
* In portfolio performance analysis of actively managed funds a benchmark return is often used.
* Queens Student Managed Fund (QSMF) is benchmarked against the FTSE 350 Index.

:::
:::{.column .small}
* The simple excess return and log excess return of an asset are then defined as: 

$$Z_t=R_t-R_{0t}, \; z_t=r_t-r_{0t}$$

* where _R~0t~_ and _r~0t~_ are the simple and log returns of the reference asset, respectively. 
* In finance literature, the excess return is thought of as the payoff on an arbitrage portfolio that goes long in an asset and short in the reference assets with no net initial investment.
* In portfolio analytics, active return is the excess return compared to the *passive* benchmark return.
:::
:::

## Examples

::: columns
:::{.column}
- Simple return Vs Log returns

```{r, echo=FALSE}
#library(fpp2)
library(tidyverse)
glen %>%
  select(date,simple_return,log_return) %>%
  drop_na() %>%
  pivot_longer(!date,names_to ="Return_Type",values_to="return") %>%
  ggplot(aes(x=date,y=return)) +
  geom_line() +
  facet_wrap(~Return_Type,ncol = 1)
  
```
:::
::: {.column.discussion}

* Plot shows a time plot of daily simple and log returns of Glencore stock from `r first(glen$date)` to `r last(glen$date)`.
* There are `r sum(!is.na(glen$log_return))` observations.
* From the plots, the behaviour of log returns is very similar to simple returns.
* The correlation coefficient between log and simple returns is 0.9997.
* This is expected, as when x is close to zero, $log(1+x) \approx x$ and daily simple returns of Glencore are small in the sampling period.

:::
:::

## Financial Returns Characteristics

* Daily log returns are simply the change series of log prices.
* In `R`, a change series can easily be obtained by taking the *difference* of log prices `diff(log(price))`.
* In Figure 1.2  the *adjusted* daily price is used as it accounts for stock splits.
* In Figure 1.1 the total return calculations includes both stock splits and dividend income.



## Inference from time plot of log returns

  1. There exist some large outlying observations
  2. The returns were volatile in certain periods but stable in others.
  We can thus infer:
  
  1. Statistically the returns have heavy tails
  2. The returns series is characterised by *volatility clustering*


## Bond yields and prices

### What is a Bond?

  * A financial instrument which pays the face value (par value) to its holder at the time of maturity.
  * Some bonds also pay interest periodically; *coupon payments*.
  * Zero-coupons bonds do not pay periodic interest.
  * Bond yield is the return recieved for holding a bond to maturity.


## Current yield{.smaller}

  * The percentage return that the annual coupon payment provides the investor

$$ 
\text{Current Yield}_t = \frac{ \text{Annual Interest paid in dollars}}{\text{Market Price of the Bond}_t} \times 100%
$$
  
* For a zero-coupon bond the yield is defined as:
  
$$
\text{Current Yield}_t=\frac{ \text{Annual Interest paid in dollars}}{\text{Market Price of the Bond}_t}^{1/k}-1 \\
\text{where }k\text{ denotes time to maturity in years}
$$
* The definition shows it is typically time dependent ($_t$ subscript) and does not include capital gains/losses.


## Yield to maturity

  * Considers the time value of money by equating the bond price to the present value of future payments.
  * YTM is a more common way to measure bond investment.
  * It's calculation is more complex and calibration is required to find an accurate approximation.

## Yield to maturity  

* Suppose that the bond holder will receive $k$ payments between purchase and maturity then:
  
$$
P=\frac{C_1}{1+y}+\frac{C_2}{(1+y)^2}+\dots+\frac{C_k+F}{(1+y)^k}
$$


* Where $P$ is the price of the bond, $y$ is YTM, $C_I$ is the ith coupon payment, and $F$ denotes face value.
* From the above equation the price of the bond has an inverse relationship to the YTM.


## US government bonds

* US government issues various bonds to finance its debts.
* These include
1. **Treasury Notes**
2. **Treasury Bonds**
3. **Treasury Bills**

## Treasury notes

* Treasury notes (T-Notes) mature in 1-10 years.
* They have a coupon payment every 6 months and a face value of $1000
* Notes are quoted on the secondary market at percantage of face value in thirty-seconds of a point.
* For example a quote of 95:08 indicates it is trading at a discount \$(95+8/32) X 1000/100 = \$953.5
* The 10-year note is the most frequently quoted in discussions of the US government bond market.

## Treasury bills

* US government issues various bonds to finance its debts.
* Treasury Bills (T-Bills) mature in one year or less.
* They pay no interest prior to maturity.
* They are sold at a *discount* of face value to create a positive yield to maturity.
* The minimum purchase is $100.
* The discount yield (%) of T-Bills is calculated via

$$
\text{Discount yield }=\frac{F-P}{F}\times\frac{360}{\text{Days till maturity}}\times \text{100}
$$

## Treasury bills

* Bills are quoted differently from notes and bonds, since bills do not pay an established interest rate.
* The return on a bill is the difference betweeen the purchase and subsequent sale price, or when held to maturity, the face value paid by the Treasury.
* Thus bills are quoted at a discount to face value, with the discount expressed as an annual rate based on a 360-day years,  for example:

|ISSUE| BID |ASK| CHANGE| YIELD|
|:--:|::|::|::|::|
|12-3-98 |5.08 | 5.06 | -.03|5.26 

## Treasury bonds

* T-Bonds have longer maturities (20-30 years).
* They have a 6 month coupon payment and are commonly issued with maturities of 30 years.


## Implied volatility

* Stock options are financial contracts which give the holder the right to buy (**call option**) or sell (**put option**) a stock at a prespecified price (**strike price**) within a given time period.
* Black-Scholes defines the price of an option as dependent on, the strike price, risk-free interest rate, current price and volatility of the stock.
* In practice, the stock volatility is not observed but backed out of the Black-Scholes formula as an *implied volatility* using calibration.

:::{.small}

#### CBOE VIX index

* The VIX index is the most well-known implied volatility.
* It measures the US stock market's expectation of 30-day volatility *implied* by averaging prices of S&P 500 index puts and calls over a wide range of strike prices.
* It is often regarded as the market fear factor, and can openly be traded using VIX futures and options.
:::

